---
title: "Detecting Grassroots Bribery and its Sources in China"
subtitle: "A Survey Experimental Approach"
author:
  - Wenfang Tang, Hong Kong University of Science and Technology
  - Yue Hu, Tsinghua University
# date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    self_contained: FALSE
    chakra: libs/remark-latest.min.js
    css: 
      - default
      - zh-CN_custom.css
      - style_ui.css
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

if (!require(pacman)) install.packages("pacman")
library(pacman)

p_load(
  tidyverse, icons, xaringanExtra, ggalt, drhutools
) 


use_xaringan_extra(c("tile_view", # O
                     "broadcast",
                     "panelset",
                     "tachyons",
                     "fit_screen"))
use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = FALSE  #<<
)


# Functions preload
set.seed(313)

theme_set(theme_minimal())
```

## Overview

+ Unboxing the grassroots bribery of China
    + Underestimated
    + Institutional reasons
&rArr; Challenge for the next stage of anti-corruption campaign

--

+ Method: National survey and experiments
    + A user-friendly way to expand the value of list-experiment results.

---

## Corruption and Anti-Corruption in China

.pull-left[
+ Waves of corruption cracking-down
+ Paths to report corruptions online and offline
+ Daily updated reports (http://www.ccdi.gov.cn/)
+ Tigers and flies
]

???

Jiang, Hu, Xi anti-corruption

Xi 2012


--


.pull-right[
<video height="400" controls>
<source src="images/grass_video.mp4">
</video>
]

???

http://news.sina.com.cn/c/2019-03-12/doc-ihsxncvh1996107.shtml

Cash and cash cards 8 times in 16 days 

---

.center[*Consequences*]

+ Regional differences of corruptions .small[(Bardhan, 2009; Bardhan, 2014; Sun and Johnston, 2009; Treisman, 2007; Rose-Ackerman and Palifka, 2016; Ramirez, 2014; Smart, 2018)]
+ Corruption, institution, informal politics, and legitimacy .small[(Ang and Jia, 2014; Kennedy and Shi, 2019; Landry, 2008; Lü, 2000; Lü, 2000; Manion, 1997; Manion, 2004; Shirk, 1993; Stromseth et al., 2017; Sun, 1991; Unger, 1991; Wang, 2013; Wang and Dickson, 2020; Zhu et al., 2013)]

.center[*Fighting*]

+ Anti-corruption vs. elite politics .small[(Jiang and Yang, 2016; Lu and Lorentzen, 2018)]
+ Anti-corruption vs. econ development .small[(Ang, 2016; Ang, 2020; Dickson, 2008; Duckett, 2001; Gong and Zhou, 2015; Huang, 2008; Mattingly, 2016; Mauro, 1995; Oi, 1992; Ong, 2012; Pei, 2016; Rithmire, 2015; Sun, 2004; Tsai, 2006; Tsai, 2015; Tsai, 2007; Tsai, 2007; Wank, 1996; Wedeman, 2004; White, 1996; Whiting, 2006; Zhu, 2017)]
+ Corruption vs. governmental anti-corruption .small[(e.g., Isa Ding, 2017)]

--

.center[Still a missing and fatal angle of corruption left]

???

What type of corruptions are we talk about?

Are these the most common types of corruptions in China? We don't know. But, more importantly, after waves, what's the most common now? 

Yuen yuen Ang (洪源远), University of Michigan, China's Gilded Age: The Paradox of Economic Boom and Vast Corruption:

Petty theft (e.g., extortion), grand theft (e.g., embezzlement)，speed theft (petty bribe for speed the administrative process faster), and access money (a lavish perks paid to powerful officials to buy exclusive, lucrative deals)

---

## Grassroots Bribery

.panelset[

.panel[.panel-name[Commonness]
.center[
Market reform + anti-corruption campaigns &rarr; Grassroots bribery .navy[&darr;]

vs

Bribery is rational + institutional role &rarr; Grassroots bribery .red[&uarr;]

&dArr;

They'll do it, but they .red[won't tell] you~ (H<sub>1</sub>)
]
]

.panel[.panel-name[Sources]
.pull-left[
*Institutional* 

Near to the center, less likely to bribe (.navy[&darr;], H<sub>2</sub>)
+ Household registration
+ Work unit ownership type
+ Level of residency
]
.pull-right[
*Non-institutional*
+ Political mobilization: TV viewers (.navy[&darr;], H<sub>3</sub>)
+ Culture: Guanxiology (interpersonal , .red[&uarr;], H<sub>4</sub>)
+ Social Economy: Long-term: Education (.navy[&darr;], H<sub>5.1</sub>); Short-term: Family income (.red[&uarr;], H<sub>5.2</sub>)
]

.center[
*Controls*    
Gender, age, multilevel modeling (intercept varying at the county level),   
multiple imputations for the missing
]
]
]



???

Regulation theory: market enhances the role of formal institutions, and governmental regulation reduces informal economic activities

Supply theory: Institutions holding the resources of educaiton, healthcare, business licensing, legal system etc. 

+ Disposable income
+ Ineffective deterrence
+ High bureaucratic control

Social desirability making things harder to detect

+ Household registration
    + Urban people practice more 
    + Migrants having more disposal money

+ work unit ownership type
    + Party-state organizations less
    
+ level of residency: The higher liers less

Migrants have more disposal money

---

## Expose Grassroots Bribery in China

2018 China Grassroots Bribery Behavior Survey

2,581 respondents 

Nationally representative

--

Public sector and grassroots bribery engagements

+ Education (37%, .red[7.12%])
+ Health care (54%, .red[8.62%])
+ Local government offices (22%, .red[6.77%])
+ Public security and law (17%, .red[6.44%])

???

Within those who interact

--

Four list experiments (direct + indirect questions)


---

## They'll Do It, But They Won't Tell You

```{r list-experiment, fig.align='center', fig.width=10}
df_bribery <- tibble::tribble(
            ~bribery, ~direct, ~experiment,
       "Health Care",    0.0712,       0.1836,
         "Education",    0.0862,        0.174,
  "Local Government",    0.0677,        0.0876,
    "Police and Law",    0.0644,       0.2499
  ) %>% 
  mutate(bribery = factor(bribery, levels = rev(c("Health Care", "Education", "Local Government", "Police and Law"))))

percent_first <- function(x) {
  x <- sprintf("%d%%", round(x*100))
  x[2:length(x)] <- sub("%$", "", x[2:length(x)])
  x
}

ggplot(data = df_bribery, aes(y = bribery)) +
  geom_segment(aes(yend = bribery,
                   x = 0,
                   xend = 0.3),
               size = 0.15) +
  geom_dumbbell(
    aes(x = direct, xend = experiment),
    size = 1.5,
    size_x = 3,
    size_xend = 3,
    colour = gb_cols("light grey"),
    colour_x = "black",
    colour_xend = gb_cols("gold")
  ) +
  geom_text(
    data = filter(df_bribery, bribery == "Health Care"),
    aes(x = direct, label = "Direct Questioning"),
    color = "black",
    size = 6,
    vjust = -1.5,
    fontface = "bold"
  ) +
  geom_text(
    data = filter(df_bribery, bribery == "Health Care"),
    aes(x = experiment, label = "List Experiment"),
    color = gb_cols("gold"),
    size = 6,
    vjust = -1.5,
    fontface = "bold"
  ) +
  geom_text(
    aes(x = direct, label = percent_first(direct)),
    color = "black",
    size = 6,
    vjust = 2.5
  ) +
  geom_text(
    aes(x = experiment, label = percent_first(experiment)),
    color = gb_cols("orange"),
    size = 6,
    vjust = 2.5
  ) +
  theme(axis.title = element_blank(), 
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 16))

```

---

## Extending the Information from the Experiments

Frequency Weight based on the .red[average intention to hide] the true answers

\begin{align}
W_i =& \sum I_i^{type}\cdot\frac{O^{type}}{t_i},\\
=& I_i^{health}\cdot\frac{O^{health}}{t_i} + I_i^{education}\cdot\frac{O^{education}}{t_i} +\\ 
&I_i^{local\  gove}\cdot\frac{O^{local\ gov}}{t_i} + I_i^{judicial}\cdot\frac{O^{judicial}}{t_i}.
\end{align}

.center[
$I_i^{type}$: 0/1, whether *i* report;

$O^{type}$: Experiment/direct;

$t_i$: Type counts that *i* reports
]

---

## Findings: Unsurprises and Surprises

+ 47.65% (weighted)

+ Judicial system vs. Local government

```{r bribery-perception, fig.align='center', fig.height=6}
df_perception <- tibble::tribble(
    ~perception, ~value,        ~type,
    "Tolerance",  0.3548,  "tolerance",
       "Safety",  0.5978,  "tolerance",
      "Central",  0.3436, "perception",
       "County",  0.5248, "perception",
  "Residential",  0.5656, "perception"
  ) %>% 
  mutate(perception = factor(perception, levels = c("Residential", "County", "Central", "Safety", "Tolerance")))


ggplot(df_perception, aes(y = perception)) +
  geom_segment(aes(yend = perception,
                   x = 0.2,
                   xend = 0.65),
               size = 0.15) +
  geom_point(aes(x = value, y = perception, color = type), size = 5) + 
  geom_text(
    aes(x = value , label = scales::percent(value)),
    color = "black",
    size = 5.5,
    vjust = 2.5
  ) + 
  scale_color_gb(reverse = TRUE) +
  theme(axis.title = element_blank(), 
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 16),
        legend.position = "none")

```

---

.center[
<img src = "images/coefplot.png" height = 600>
]

???

Surprises:

Non-profit units most corrupted:

1. 公共部门或政府
2. 私营部门
3. 非盈利组织
4. 个体
8. 不好说
9. 拒绝回答

City type: Not the top or the bottom, but in middle, necessary recources + less supervision

Television: Learning the clue

---

## Implications: When the Missing Angle Is Encovered

+ How common
    + A step further to uncover the reality

--

+ Sources: Institutional + noninstitutional

--

+ Where does anticorruption go
    + Focusing on the source of the problem
    + Understanding the underlying logic

---

class: inverse, center, middle

# Thank you!

`r feather_icons("mail")`&nbsp;[yuehu@tsinghua.edu.cn](mailto:yuehu@tsinghua.edu.cn) 

`r feather_icons("globe")`&nbsp; https://sammo3182.github.io/

`r feather_icons("github")`&nbsp; [sammo3182](https://github.com/sammo3182)